---
title: "Using excursions with INLA"
author: "David Bolin and Finn Lindgren"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using excursions with INLA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ref.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(excursions)
library(RColorBrewer)
library(INLA)
library(splancs)
set.seed(123)
```

```{r inla_link, include = FALSE}
inla_link <- function() {
  sprintf("[%s](%s)", "`R-INLA`", "https://www.r-inla.org")
}
```

## Introduction

The function `excursions.inla` is used to compute excursion sets and credible regions for contour curves for latent Gaussian models that have been estimated using `INLA`. It takes the same arguments `excursions`, except that `mu` and `Q` are replaced with arguments related to `INLA`. A basic call to the function `excursions.inla` looks like
```{r, eval=FALSE}
excursions.inla(result.inla, name, alpha, u, type)
```
Here `result.inla` is the output from a call to the `inla` function, and `name` is the name of the component in the output that the computations should be done for. For more complicated models, one typically specifies the model in `INLA` using an `inla.stack` object. If this is done, the call to `excursions.inla` will instead look like
```{r,eval=FALSE}
excursions.inla(result.inla, stack, tag, alpha, u, type)
```
Here `stack` is now the stack object and `tag` is the name of the component in the stack for which the computations should be done. The typical usage of the `tag` argument is to have one part of the stack that contains the locations where measurements are taken, and another that contains the locations where the output should be computed.

`excursions.inla` has support for all strategies described in [Definitions and computational methodology](theory.html) for handling latent Gaussian structures: The argument `method` can be one of `'EB'`, `'QC'`, `'NI'`, `'NIQC'`, and `'iNIQC'`.

Similarly `contourmap.inla` and `simconf.inla` are the `INLA` versions of the `contourmap` and `simconf` functions. The model specification using these functions is identical to that in the `excursions.inla` function. We now provide two applications to illustrate the methods of the package using the `INLA` interface.

### An example using a simultaneous confidence bands

We start by a temporal example to illustrate how a simultaneous confidence band can be created using `simconf.inla`. We use the much analyzed binomial time series from [@kitagawa1987non]. Each day during the years 1983 and 1984, it was recorded whether there was more than 1 mm rainfall in Tokyo. Of interest is to study the underlying probability of rainfall as a function of day of the year. The data is modelled as $y_i \sim Bin(n_i,p_i)$ for calendar day $i = 1,\dots,366$. Here $n_i = 2$ for all days except for February 29 ($i=60$) which only occurred during the leap year of 1984. The probability $p_i$ is modeled as a logit-transformed Gaussian process.

The model and the following `INLA` implementation of the model is described further in [@lindgren2015software].
```{r}
data("Tokyo")
mesh <- inla.mesh.1d(seq(1, 367, length = 25), interval = c(1, 367),
                     degree = 2, boundary = "cyclic")
kappa <- 1e-3
tau <- 1 / (4 * kappa^3)^0.5
spde <- inla.spde2.matern(mesh, constr = FALSE, 
                          theta.prior.prec = 1e-4,
                          B.tau = cbind(log(tau), 1), 
                          B.kappa = cbind(log(kappa), 0))
A <- inla.spde.make.A(mesh, loc = Tokyo$time)
time.index <- inla.spde.make.index("time", n.spde = spde$n.spde)
stack <- inla.stack(data = list(y = Tokyo$y, link = 1, Ntrials = Tokyo$n),
                    A = list(A), effects = list(time.index), tag = "est")
data <- inla.stack.data(stack)
```
Next, the model is estimated using the `inla` function. Since we want to analyze the output using `excursions`, the additional option `control.compute = list(config = TRUE)` must be specified in the `inla` function. This makes the function save some extra output needed by `excursions`.
```{r}
result <- inla(y ~ -1 + f(time, model = spde), 
               family = "binomial", 
               data = data,
               Ntrials = data$Ntrials, 
               control.predictor = list(A = inla.stack.A(stack), 
                                        link = data$link, 
                                        compute = TRUE),
               control.compute = list(config = TRUE),
               num.threads = "1:1")
```
We now have estimates of the posterior mean and marginal confidence intervals for the probability of rain for each day. However, if we also want joint confidence bands, we can estimate these using `simconf.inla` as
```{r}
res <- simconf.inla(result, stack, tag = "est", alpha = 0.05, link = TRUE,
                    max.threads = 2)
```
Note the argument `link=TRUE` which tells the function that the results should be returned in the scale of the data, and not in the scale of the linear predictor. Next, we plot the results, showing the marginal confidence bands with dashed lines and the simultaneous confidence band with dotted lines. 
```{r}
index <- inla.stack.index(stack, "est")$data
plot(Tokyo$time, Tokyo$y / Tokyo$n, xlab = "Day", ylab = "Probability")
lines(result$summary.fitted.values$mean[index])
matplot(cbind(res$a.marginal, res$b.marginal), type = "l", lty = 2,
        col = 1, add = TRUE)
matplot(cbind(res$a, res$b), type = "l", lty = 3, col = 2, add = TRUE)
```

### A spatial example 

We will now use a spatial dataset to illustrate how `excursions` can be used. In order to keep the parts of the code that are not relevant to `excursions` brief, we again use data available in `INLA`. The dataset consists of daily rainfall data for each day of the year 2011, at 616 locations in and around the stated of Paran\'a in Brazil. In the following analysis, we analyze the data from the month of January.

The statistical model used for the data is a latent Gaussian model, where the precipitation measurements are assumed to be $\Gamma$-distributed with a spatially varying mean. The mean is modeled as a log-Gaussian Mat\'ern field specified as an SPDE model. Details of the current model model and the following INLA implementation can be found in the SPDE tutorial available on the `INLA` homepage, see [@Wallin15] for an analysis of the data using a different non-Gaussian SPDE model.

We start by loading the data and defining the model:
```{r}
data("PRprec")
data("PRborder")
Y <- rowMeans(PRprec[, 3 + (1 : 31)])
ind <- !is.na(Y)
Y <- Y[ind]
coords <- as.matrix(PRprec[ind, 1 : 2])
b <- inla.nonconvex.hull(coords, -0.03, -0.05, resolution = c(100, 100))
prmesh <- inla.mesh.2d(boundary = b, max.edge = c(.45, 1), cutoff = 0.2)
A <- inla.spde.make.A(prmesh, loc = coords)
spde <- inla.spde2.matern(prmesh, alpha = 2)
mesh.index <- inla.spde.make.index(name = "field", n.spde = spde$n.spde)
```
The measurement stations are spatially irregular, but we are interested in making predictions to a regular lattice within the state. In order to do continuous domain interpretations, we define the lattice locations as a lattice object using the `submesh` function of the `excursions` package. The function `inout` from the package `splancs` is used to find the locations on the lattice that are within the region of interest.
```{r}
nxy <- c(50, 50)
projgrid <- inla.mesh.projector(prmesh, xlim = range(PRborder[, 1]),
                                 ylim = range(PRborder[, 2]), dims = nxy)
xy.in <- inout(projgrid$lattice$loc, cbind(PRborder[, 1], PRborder[, 2]))
submesh <- submesh.grid(matrix(xy.in, nxy[1], nxy[2]),
                        list(loc = projgrid$lattice$loc, dims = nxy))
```
We now define the `stack` objects and estimate the model using `inla`. Again note that we have to set the `control.compute` argument of the results is to be used by `excursions`.
```{r}
A.prd <- inla.spde.make.A(prmesh, loc = submesh$loc)
stk.prd <- inla.stack(data = list(y = NA), A = list(A.prd, 1),
                      effects = list(c(mesh.index, list(Intercept = 1)),
                                     list(lat = submesh$loc[,2], 
                                          lon = submesh$loc[,1])), 
                      tag = "prd")
stk.dat <- inla.stack(data = list(y = Y), A = list(A, 1),
                      effects = list(c(mesh.index, list(Intercept = 1)),
                                     list(lat = coords[,2],
                                          lon = coords[,1])), 
                      tag = "est")
stk <- inla.stack(stk.dat, stk.prd)
r <- inla(y ~ -1 + Intercept + f(field, model = spde), 
          family = "Gamma",
          data = inla.stack.data(stk), 
          control.compute = list(config = TRUE,
                                 return.marginals.predictor=TRUE),
          control.predictor = list(A = inla.stack.A(stk), 
                                   compute = TRUE,
                                   link = 1),
          num.threads = "1:1")
```
We now want to find areas that likely experienced large amounts of precipitation. In the following code, we compute the excursion set for the posterior mean for the level $7$ mm of precipitation. To indicate that this level is in the scale of the data, and not in the scale of the linear predictor, we use the `u.link=TRUE` argument in the `excursions` call.
```{r}
exc <- excursions.inla(r, stk, tag = "prd", u = 7, u.link = TRUE,
                      type = ">", F.limit = 0.6, method = 'QC',
                      max.threads = 2)
sets <- continuous(exc, submesh, alpha = 0.1)
```
We also compute the contour curve for the level of interest on the continuous domain, using the `tricontourmap` function.
```{r}
con <- tricontourmap(submesh, z = exc$mean, levels = log(7))
```
We plot the resulting continuous domain excursion function together with the contour curve using the following commands. 
```{r}
cmap.F <- colorRampPalette(brewer.pal(9, "Greens"))(100)
proj <- inla.mesh.projector(sets$F.geometry, dims = c(300,200))
image(proj$x, proj$y, inla.mesh.project(proj, field = sets$F),
      col = cmap.F, axes = FALSE, xlab = "", ylab = "", asp = 1)
plot(con$map, add = TRUE)
```

To visualize the posterior mean using a contour map using the following commands
```{r}
cmap <- colorRampPalette(brewer.pal(9, "YlGnBu"))(100)
lp <- contourmap.inla(r, stack = stk, tag = "prd", n.levels = 2,
                      compute = list(F = FALSE))
tmap <- tricontourmap(submesh, z = lp$meta$mu, levels = lp$meta$levels)
plot(tmap$map, col = contourmap.colors(lp, col = cmap))
```

Here, the `contourmap.inla` computes the levels of the contour map and `tricontourmap` computes the contour map on the mesh. Finally, `contourmap.colors` is used to compute appropriate colours for visualizing the contour map. 

The contour map we computed had two contours, and a relevant question is now if this is an appropriate number. To investigate this, we compute the $P_2$ quality measure for this contour map and for contour maps with one and three levels.
```{r}
lps <- list()
for(i in 1:4) {
    lps[[i]] <- contourmap.inla(r, stack = stk, tag = "prd", n.levels = i,
                                compute = list(F = FALSE, measures = c("P2")),
                                max.threads = 2)    
}
print(data.frame(P2 = c(lps[[1]]$P2, lps[[2]]$P2, lps[[3]]$P2, lps[[4]]$P2),
                 row.names = c("n.level = 1","n.level = 2",
                               "n.level = 3","n.level = 4")))

```
We see that using only one or two contours give very high credibility, whereas using four contours give a credibility that is close to zero. The contour map with three contours has a credibility around $0.2$ and seems to be a good compromise for this application.

## References

